Characterizing narrative entities with embeddings : a case study of character relationships in Les Misérables

### 2022-10-11 article/paper1.tex
write paper

Note that vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ obtained from \textbf{CA} reflect textual unit profile (in terms of words) regarding the mean profile (the origin in the factorial map). This analysis is thus \emph{contrastive}: it highlights unit variations among the studied text. This could becomes problematic in order to study a dataset containing multiple corpora: units belonging to the same text will have a greater chance to be close from each other, and the displayed variance might be largely composed by differences of style between the different works. Another limitation with this approach is that the words helping the interpretation of units (and entities) are contained in the studied text. Approaches requiring to study the position of units and entities relatively to a predefined list of words (e.g., friends, enemies, family) might therefore be impossible if these words do not appear in the text.

@@

The five volumes of \emph{Les Misérables}, in French, were extracted from \emph{Project Gutenberg}\footnote{\url{https://www.gutenberg.org/}}, while headers and footers of each files were manually removed. The whole text was lower cased, lemmatized, and stopwords\footnote{from a list made by Jacques Savoy \url{http://members.unine.ch/jacques.savoy/clef/frenchST.txt}} and punctuation were removed. Volumes, books, and chapters breaking points were kept for later uses.

@@

Characters were detected using \emph{Flair}\footnote{\url{https://github.com/flairNLP/flair}} (REF) NER tools. In order to unify character and to further refine the detected characters list, we used hand-made lists of character and aliases from NER results. It resulted in the detection of 54 characters. The entities considered in table $\mathbf{E}$ (Figure \ref{occ_table}) are single character (54 entities) and character pairs (547 entities), resulting in a table of size $365 \times 601$. A character or a character-pairs presence is considered present iff characters are detected at least 2 times in the chapter.

@@

The \textbf{CA} method is fully automated, and results in vectors in a $364$-dimensional space, while the $\textbf{WV}$ methods is based on pre-trained word vectors using \emph{fastText} (REF) trained on Common Crawl.\footnote{\url{ https://fasttext.cc/docs/en/crawl-vectors.html}, accessed September 2022.} For French, the number of word-vectors $m$ is 2 millions and the dimension of vectors are $300$.

Note that, in addition to having two table 4 variations for the methods, 2 tables $\mathbf{E}$ and $\mathbf{E}_\text{vol}$, and a considerable number of entities, results can also be presented in various ways (similarities between entities, between entities and words, etc.). Thus, we chose to show here a selection of results for the each method: the 5 most associated words regarding entities (section \ref{words}), the 5 most associated entities regarding a subset of words (section \ref{objects}), and a diachronic study of top associated words for a subset of entities (section \ref{diachronic}). We invite curious readers to consult results for all words and entities, which can be found in our GitHub repository.\footnote{\url{https://github.com/gguex/char2char_vectors/results}}

@@

These results are extracted from the transposed table from the last, and display the most associated entities to a selected set of words. They can be found in Table 4. This type of results can be seen as queries, made from a single word, which output the most associated entities in the work related to that query. We chose here to show top entities related to words "aimer", "rue", "justice" and "guerre", as they represent some of the main topics of the book. In this task again, from our point of view, the \textbf{CA-REG} displays the most accurate results: the main love relationship (Cosette-Marius) of the book is the most associated entity for "aimer", several "amis de l'ABC" (a revolutionary group) are most associated with "rue", the cop-suspect relationship (Javert-Valjean) is the top entity for "justice", and military officers or bellicose characters are associated with "guerre". While somewhat inferior with the selected set of queries, \textbf{WV} methods have the advantage to be able to query words outside the scope of the book, as the pre-trained word embedding possess a very large vocabulary. It should also give interesting results when comparing different books, due to the fixed nature of word-vectors, but case studies concerning multiple works are outside the scope of this article.

@@

	\caption{The 5 most associated words (similarity in parentheses) vs volumes constant, Valjen, Cosette, and Cosette-Valjean entity, as found in each tome regarding the \textbf{CA-REG} and \textbf{WV-REG} methods ($\lambda = 0.01$).}

@@

These results are obtained from the table $\mathbf{O}_\text{vol}$ where entities are considered different based on the volume. By doing so, it permits to track the evolution of similarities along the book. Additionally to entities, we can also define constant term $V_i$ for each volumes $i$, which absorbs the associated words with each volume. Results for constants and a subset of entities (Valjean, Cosette, Valjean-Cosette) can be found in Table 5. Note that we did not show \textbf{CONT} results in this table, as they are similar to the one found in Table 3 : words are often repeated themselves for different objects and are less convincing.

Here again, we see that associated words for the \textbf{WV} give a general tone to tomes and entities, while \textbf{CA} results are more specific and related to particular events which occurred for characters. As expected, volume constants give a quick summary of each volumes, especially with the \textbf{CA-REG} method (e.g. $V_2$ for the Battle of Waterloo, $V_4$ for the barricade event). Associated words with entities also seems accurate in describing them. Note that Cosette was not detected in volume 3 because she is not explicitly cited (she is often refereed as "the daugther of M. Leblanc"), and this  also entails an absence of the Valjean-Cosette pair.

@@

>>>>

Note that vectors $\mathbf{x}_1, \ldots, \mathbf{x}_n$ obtained from \textbf{CA} reflect textual unit profile (in terms of words) regarding the mean profile (the origin in the factorial map). This analysis is thus \emph{contrastive}: it highlights unit variations among the studied text. It means that the particular tone of the whole studied text might be hidden in this analysis and only the variation around this tone will be revealed. It might lead to the situation where the (absolute) feeling experienced by the reader will not appear in this analysis, e.g. a sad character in a sad book will appear joyful if he is less sad than the mean tone. This could becomes problematic when this method is used sequentially to study multiple works: particularities of each book will be hidden. Another limitation with this approach is that the words helping the interpretation of units (and entities) are contained in the studied text. Approaches requiring to study the position of units and entities relatively to a predefined list of words (e.g., friends, enemies, family) might therefore be impossible if these words do not appear in the text.

@@

The five volumes of \emph{Les Misérables}, in French, were extracted from \emph{Project Gutenberg}\footnote{\url{https://www.gutenberg.org/}.}, while headers and footers of each files were manually removed. The whole text was lower cased, lemmatized, and stopwords\footnote{from a list made by Jacques Savoy \url{http://members.unine.ch/jacques.savoy/clef/frenchST.txt}.} and punctuation were removed. Volumes, books, and chapters breaking points were kept for later uses.

@@

Characters were detected using \emph{Flair}\footnote{\url{https://github.com/flairNLP/flair}.} (REF) NER tools. In order to unify character and to further refine the detected characters list, we used hand-made lists of character and aliases from NER results. It resulted in the detection of 54 characters. The entities considered in table $\mathbf{E}$ (Figure \ref{occ_table}) are single character (54 entities) and character pairs (547 entities), resulting in a table of size $365 \times 601$. A character or a character-pairs presence is considered present iff characters are detected at least 2 times in the chapter.

@@

The \textbf{CA} method is fully automated, and results in vectors in a $364$-dimensional space, while the $\textbf{WV}$ methods is based on pre-trained word vectors using \emph{fastText} (REF) trained on Common Crawl.\footnote{\url{ https://fasttext.cc/docs/en/crawl-vectors.html}.} For French, the number of word-vectors $m$ is 2 millions and the dimension of the vector space is $300$.

Note that, in addition to having two table 4 variations for the methods, 2 tables $\mathbf{E}$ and $\mathbf{E}_\text{vol}$, and a considerable number of entities, results can also be presented in various ways (similarities between entities, between entities and words, etc.). Thus, we chose to show here a selection of results for the each method: the 5 most associated words regarding entities (section \ref{words}), the 5 most associated entities regarding a subset of words (section \ref{objects}), and a diachronic study of top associated words for a subset of entities (section \ref{diachronic}). We invite curious readers to consult results for all words and entities, which can be found in our GitHub repository.\footnote{the "results" folder in \url{https://github.com/gguex/char2char_vectors}.}

@@

These results are extracted from the transposed table from the last, and display the most associated entities to a selected set of words. They can be found in Table 4. This type of results can be seen as queries, made from a single word, which output the most associated entities in the work related to that query. We chose here to show top entities related to words "aimer", "rue", "justice" and "guerre", as they represent some of the main topics of the book. In this task again, from our point of view, the \textbf{CA-REG} displays the most accurate results: the main love relationship (Cosette-Marius) of the book is the most associated entity for "aimer", several "amis de l'ABC" (a revolutionary group) are most associated with "rue", the cop-suspect relationship (Javert-Valjean) is the top entity for "justice", and military officers or bellicose characters are associated with "guerre". While somewhat inferior with the selected set of queries, \textbf{WV} methods have the advantage to be able to query words outside the scope of the book, as the pre-trained word embedding possess a very large vocabulary.

@@

	\caption{The 5 most associated words (similarity in parentheses) vs volumes constant, Valjen, Cosette, and Cosette-Valjean, as found in each tome regarding the \textbf{CA-REG} and \textbf{WV-REG} methods ($\lambda = 0.01$).}

@@

These results are obtained from the table $\mathbf{E}_\text{vol}$ where entities are considered different based on the volume. By doing so, it permits to track the evolution of similarities along the book. Additionally to entities, we can also define constant term $V_i$ for each volumes $i$, which absorbs the associated words with each volume. Results for constants and a subset of entities (Valjean, Cosette, Cosette-Valjean) can be found in Table 5. Note that we did not show \textbf{CONT} results in this table, as they are similar to the one found in Table 3 : words are often repeated themselves for different objects and are less convincing.

Here again, we see that associated words for the \textbf{WV} give a general tone to tomes and entities, while \textbf{CA} results are more specific and related to particular events which occurred for characters. As expected, volume constants give a quick summary of each volumes, especially with the \textbf{CA-REG} method (e.g. $V_2$ for the Battle of Waterloo, $V_4$ for the barricade event). Associated words with entities also seems accurate in describing them. Note that Cosette was not detected in volume 3 because she is not explicitly cited (she is often refereed as "the daugther of M. Leblanc"), and this also explains the absence of the Cosette-Valjean pair.

@@

In this article, we introduced a general framework in order to automatically extract textual information about narrative entities from a small corpus or a single work. The framework is build on two tables, the unit-word table $\mathbf{N}$ and the unit-entity table $\mathbf{E}$. This data organization sets subsequent analyses into a classical statistical framework, where the goal is to see how variables in $\mathbf{E}$ (the entities) affects the variables in $\mathbf{N}$ (the vocabulary) for each textual units. A choice was taken to use embeddings for analyzing these effects: units and words are embedded using Correspondence Analysis or pre-trained Word Embedding on $\mathbf{N}$, and entities are embedded in the same space as units using the Centroids or the Regressions methods on $\mathbf{E}$. These embeddings are then used in order to see affinities between entities and words, enabling the characterization of the formers by the latters. A case study on \emph{Les Misérables} was performed to see if methods gave promising results.

The first important choice in the analysis is how to define the size of units. Other corpora were also tested (e.g. Shakespeare plays) and it seems important to define units with at least a paragraph size (after preprocessing) in order to represent them accurately. Choosing small units might be able to successfully capture word specificities related to a small subset of entities, but their direction becomes almost orthogonal one to another if the size of units is too small. This situation results in taking into account the "noise" in subsequent analysis, typically denoting an overfitting regime with an high variance and low bias. By contrast, large units will result in an underfitting regime, with a low variance and high bias, failing to capture entities specificities, but more robust to particularities in word usages. Having enough units is also important in order to properly locate entities in the embedding space. In order to analyze characters and relationships, we advice the practitioner to use his prior knowledge of the work in order to split the studied narrative as close as possible to "scenes" (as found in theater), which describes a particular event between an almost constant set of characters.

The second choice is to select which entities to study. This choice is of course driven by the studied problematic, but is also limited by the automatic extraction tools available. These entities can be quite various, but must appear frequently in the work in order to be placed correctly in the embedding space. However, it is unadvised to set an entity which is almost always there (e.g. a narrator), as it will already be represented by the origin in the $\mathbf{CENT}$ method or as the constant term in the $\mathbf{REG}$ method. As a rules of thumbs, the number of entities should ideally be lower than the number of textual units. However, even with an exceeding number of entities (like in our case study, where we had 547 entities for 365 units), if some entities appear rarely, analyses are still possible. Note that the version of the table $\mathbf{E}$ containing counts of entities rather than presence among each units was also tested in experiments, but gave similar results for the studied corpus.

The choice of using embeddings, where units, entities and words are located, is motivated by the fact that the resulting space permits many types of exploration. As presented in this article, we can extract some of the most (or least) associated words with each entity or rank entities according to a word query. Nevertheless, other types of exploration could be made. Entities could be placed along a particular axis in the space, define with a set of positive and a set of negative words in order to highlight a particular contrast (good-bad, like in sentiment analysis, introvert-extrovert, friend-enemy, etc.). This approach could also be combined with a clustering of the words, or a Topic Modeling methods, thus permitting to define interpretable regions of the embedding space. Relative location of entities could also be used in order to cluster or classify them. All these leads could be explored by subsequent researches.

The difference in the choice between \textbf{CA} and \textbf{WV} embeddings appears quite clearly in the results. \textbf{CA} highlights particular words associated with entities, very specific to the studied work and the narrative events found in it, while \textbf{WV} give a feel of the general tone in the text when these entities are present. This difference is explained by the fact that \textbf{CA} focuses on words appearing within the work, with possibly very different location to semantically similar words, while \textbf{WV} word-vectors are positioned regarding their semantic and syntactic similarities. A entity located in the \textbf{WV} space will then be in a semantic or syntactic region, and its characterizing words should all be related. Results show that \textbf{CA} methods generally perform better to quickly interpret entities among the narrative, but might bit limited for some applications. As a matter of fact, the advantage of \textbf{WV} embedding is that its space is absolute, permitting the comparison of results between sequentially studied texts, and also contains a larger vocabulary in its embedding. This last property could be used in order to use a fixed list of relationship attribute (e.g., friend, enemy, family, colleague), which do not necessarily appear in every text in the studied corpus, in order to categorized character relationships.

The choice between the \textbf{CENT} method and the \textbf{REG} method is relatively easy: thanks to its hyperparameter $\lambda$, the \textbf{REG} method can give similar results to the \textbf{CENT} method when $\lambda$ is high (with only a contraction of entity vectors), but also give more "perpendicular" word affinities between entities when $\lambda$ is low. Thus, it is clearly a superior choice in order to give a variety of results. The choice for this hyperparameter $\lambda$ depends on what the practitioner want from his results. If her/his entities are defined that some of them are completely included into other, such as character and character-pair, and she/he would like to have specificities about the finer grained entity, the $\lambda$ must be set to a low value. By contrast, if he do not mind to have some of her/his entities defined as a mixture of others, he can set $\lambda$ to a high value. However, very low values of $\lambda$ should be avoided if the number of entities is high compared to the number of units, as this will lead to an overfitting of regression coefficients and will result in the association of very rare and specific words with entities.

Finally, the biggest weakness of this framework yet is the difficulty to validate its pertinence. Several other case studies, with results carefully scrutinized regarding prior knowledge, should be undertaken in order to see if it result are solid, but this type of experiments are expansive both in time and explanation length. Another idea could be to use an annotated corpora such as in (REF), where human annotators already classified character relationships in various categories, and to see if the presented method can actually retrieve the different categories. Such an experiment can be promising, but it currently need highly performative automatic tools in order to detect and unify character entities occurrences, which is currently missing. Nevertheless, first case studies gave promising results for this framework, and its flexibility could lead to various applications.
